MCFA on Galah

### 2018-10-31 article/ms.tex
some methods

\documentclass[twocolumn]{aastex61}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{color}
\usepackage{comment}

@@

@@

\section{Model} \label{sec:model}

>>>>

\documentclass[twocolumn]{aastex61}
\usepackage{bm}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{color}
\usepackage{comment}

@@

\newcommand{\vect}[1]{\boldsymbol{\mathbf{#1}}}
\renewcommand{\vec}[1]{\vect{#1}}

\newcommand{\weight}{\pi}
\newcommand{\data}{\textbf{Y}}
\newcommand{\vecdata}{\vec\data}

\newcommand{\nextstep}{^\textrm{(t+1)}}
\newcommand{\thisstep}{^\textrm{(t)}}
\newcommand{\transpose}{^\intercal}
\newcommand{\eye}{\textbf{I}}

\newcommand{\factorloads}{\textbf{L}}
\newcommand{\factorscores}{\textbf{S}}

\newcommand{\scoremeans}{\vec\xi}
\newcommand{\scorecovs}{\vec\Omega}

@@

\section{Methods} \label{sec:methods}

% Describe how we will attempt to define a generative model for the data, from the latent space.

In this work the data $\vecdata$ is a $n \times d$ matrix 
where $n$ is the number of stars and $d$ is the number of
chemical abundances measured for each star. We assume a
generative model for the data 
\begin{equation}
	\vecdata = \factorloads\factorscores + \vec{e}
	\label{eq:generative-model}
\end{equation}

\noindent{}where $\factorloads$ is a $p \times d$ matrix of factor
loads that is common to all data points, and the factor scores
\begin{equation}
	\factorscores \sim \mathcal{N}(\vec\xi_k, \vec\Omega_k)
\end{equation}
\noindent{}is a \todo{$n \times p$} matrix where the entries
are drawn from $k$ multivariate normal distributions.
We assume $\vec{e} \sim \mathcal{N}\left(\vec{0}, \vec\psi\right)$
is independent of the latent space, and $\vec\psi$ is a
diagonal matrix of $d$ entries.

In this model each data point can be represented as being drawn
from a mixture of multivariate normal components, except the components
are \emph{clustered in the latent space} $\factorscores$ and projected
into the data space by the factor loads $\factorloads$. 
We assume that the latent space is lower dimensionality than the
data space (e.g., $p < d$).
Within the context of stellar abundances, the factor loads
$\factorloads$ can be thought of as the \emph{mean} yields
of nucleosynthetic
events (e.g., $s$-process yields from AGB stars), and the
factor scores are analogous to the relative counts of those 
nucleosynthetic events. The clustering in factor scores
achieves the same as a clustering procedure in data space,
except we simultaneously estimate the factor loads
(nucleosynthetic yields) that are common to all data points 
(stars). Within this framework a rare nucleosynthetic event
can still be described as a `common factor load' $\factorloads_i$, 
but its rarity would be represented by associated factor
scores being zero for most stars and thus have no contribution
to the observed abundances.

Without loss of generality the density of the data $\vecdata$ can be modelled as
\begin{equation}
	f(\textbf{y}; \vec\Psi) = \sum_{k=1}^{K}\weight_{k}\phi(\textbf{y};\factorloads\scoremeans_k, \factorloads\scorecovs_k\factorloads\transpose + \eye\vec\psi)
\end{equation}
\noindent{}given $p$ common factor loadings and $K$ components
clustered in the latent (factor score) space. Here the parameter
vector
$\vec\Psi$ includes $\{\factorloads,\vec\pi,\scoremeans,\scorecovs,\vec\psi\}$, and $\phi(\textbf{y};\vec\mu, \vec\Sigma)$
describes the density of a multivariate gaussian distribution with
mean $\vec\mu$ and covariance matrix $\vec\Sigma$,
and $\weight_k$ describes the relative weighting of the $k$th
component in latent space and $\sum\weight_k = 1$.
The log likelihood is then given by
\begin{equation}
	\log\mathcal{L}(\vecdata|\vec\Psi) = \sum_{k=1}^{K}\log{f(\vecdata;\vec\Psi)} \quad .
\end{equation}

The model described by Equation~\ref{eq:generative-model} is indeterminate:
there is no unique solution for the factor loads $\factorloads$ and scores
$\factorscores$. However, an accurate estimate of the parameter vector $\vec\Psi$ can be obtained by the expectation-maximization algorithm \citep{EM}.

\subsection{Initialisation}

We use the K-means++ algorithm to initially assign data points
randomly to $K$ clusters (in data space). This provides a so-called
hard initial association where each data point is wholly assigned
to a single component. Specifically we calculate the $n \times K$
responsibility matrix $\vec\tau$, where a entry of 1 indicates the
$n$th data point is associated with the $k$th cluster, and an
entry of 0 indicates no association such that
\begin{equation}
	\weight_k = \frac{1}{N}\sum_{n=1}^{N}\tau_{nk} \quad .
\end{equation}
We then chose 1,000
random data points and factorize that matrix using singular-value
decomposition. We take the principal $p$ eigenvectors as the 
initial guess for the factor loads $\factorloads$, and calculate
an initial estimate of the factor score means
\begin{equation}
	\scoremeans_k = \frac{1}{N}\sum_{n=1}^{N}\tau_{nk}\vecdata_{n}\factorloads
\end{equation}
\noindent{}and the covariance matrices of the clusters in latent space
\begin{equation}
	\scorecovs_k = \textrm{cov}{\left(\tau_{nk}\vecdata_{n}\right)} \quad .
\end{equation}

\subsection{Expectation-Maximization}

At the expectation step we evaluate the log likelihood
given the model parameters $\vec\Psi$, and we re-calculate the responsibility
matrix whose entries are the posterior probability
that the $n$th data point is associated to the $k$th component,
given the data $\vecdata$ and the current estimate of the 
parameter vector $\vec\Psi$:
\begin{equation}
	\tau_{nk} = \frac{\weight_k\phi(\vecdata_n;\factorloads\scoremeans_k, \factorloads\scorecovs_k\factorloads\transpose + \eye\vec\psi)}{\sum_{g=1}^{G}\weight_g\phi(\vecdata_n;\factorloads\scoremeans_g, \factorloads\scorecovs_g\factorloads\transpose + \eye\vec\psi)} \quad .
\end{equation}

% SVD complexity is O(n^3)
% K-means++ complexity is O(log(k))

% Discuss connection to PCA, FA, MFA.

% Include macros for \vec, \data, etc.
At the maximization step we update our estimates of the parameters,
conditioned on the data $y$ and the responsibility matrix $\tau$.

\begin{equation}
	\weight_k\nextstep = \frac{1}{N} \sum_{n=1}^{N}\tau_{nk}
	% Include proof.
\end{equation}

\begin{eqnarray}
	\vec\mu_{k}\nextstep = \vec\mu_{k}\thisstep + \vec\gamma\transpose \frac{1}{N\weight_k}\sum_{n=1}^{N}{(\vecdata\transpose - \vec{}A\vec\mu_{k}\thisstep)\vec\tau_{k}}
\end{eqnarray}

% The intermediate stpe for phi:
\begin{equation}
	\vec\phi = N\sum_{k=1}^{K}\weight_k\left[\vec\Omega_k\nextstep + \vec\mu_k\nextstep(\vec\mu_k\nextstep)\transpose\right]
\end{equation}

% And the update of the psi.
\begin{equation}
	\vec\psi\nextstep = \frac{1}{N}\left[\sum_{k=1}^{K}\vec\data^2\vec\tau_{k} - (\vec{}A\nextstep \vec\phi)\vec{}A\right]
\end{equation}

\section{Experiments} \label{sec:experiments}

\subsection{Toy model with generated data} \label{sec:experiment-toy-model}

\subsection{GALAH} \label{sec:experiment-galah}
